maxent model plus plot creation

### 2020-05-27 src/Public-maxent.Rmd
review edit fiddling

  
```{r pre-install & load packages, include=FALSE}
options(java.parameters = "-Xmx2g" )#needs to run before dismo loaded - boosts memory for java to 1g
packages <- c("dismo", "stringr", "rJava", "raster", "ecospat", "bossMaps", "lattice", "tidyr", "enmSdm", "ggplot2", "plyr", "reshape2", "tidyverse", "patchwork", "purrr", "Hmisc")
source("../src/ipak.R")
ipak(packages)
```

#need to reclassify cells outside the mcp to NA for predictions
```{r}
outsidemcp <- shapefile("../data/env/landp/outside-mcp-wgs84.shp") #made in arcgis - in aea

ascdirs <- list.dirs("../data/env/ascii-original", full.names = FALSE, recursive = FALSE)

for (ad in 1:length(ascdirs)){
  timeslice <- ascdirs[ad]
  ascfiles <- list.files(paste0("../data/env/ascii-original/", timeslice), pattern = "\\.asc$", full.names = TRUE)
  for (file in 1:length(ascfiles)){
    ras<- raster(ascfiles[file])
    crs(ras) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0 "
    ascname <- str_sub(ascfiles[file], start = 38)
    ras[outsidemcp] <- 500000
    ras[ras == 500000] <- NA
    writeRaster(ras, paste0("../data/env/asciitrim/", ascdirs[ad], "/", ascname), overwrite=TRUE)
  }
}

load the presence-background dataset 
("../output/bio/presab_nafo_yr.csv")

select month
```{r month of interest -change}
mthna <- "dec"
mthno <- "12"
```

```{r load pres-background dataset}
presback <- read.csv("../output/bio/presab_nafo_yr.csv", header = TRUE)
colnames(presback)
```

remove all rows with the year 2015 (due to low number of samples)
```{r}
presback <- subset(presback, year!="2015")
```

this dataset contains a lot of columns you don't need for the maxent models. Remove them here

```{r remove unnecessary columns}
presback <- subset(presback, select = -c(cell_id, id, decimalLatitude, decimalLongitude, datecollected, institutioncode, individualcount, depth, resname, originalscientificname, collectioncode, day, nafo_zone, gear, longitude_meters, latitude_meters, total_cell_obs_xy, total_cell_obs_xyt, bottom_depth, XXtotal_cell_obs_xyzt, temp_celsius_depth, temp_celsius_surface, longitude_meters.1, latitude_meters.1, bottom_depth_glorys, longitude_meters.2, latitude_meters.2, cell_id_3d, total_cell_obs_xyzt, cell_id_xyzt))
colnames(presback)
```

monthly model data selection and prep
For each monthly model, subset the data to the month you want, and the variables you want (Based on spearmans and VIF)

@@

@@

all layers need to be spit in bash + poweshell
- pre-split.ps1 
- separate.sh
- aftersplit.ps1

# average the maxent results

@@

@@

```{r}
library(stringr)

@@

#plotting

need some additional packages installed
```{r pre-install & load packages, include=FALSE}
packages <- c("stringr", "reshape2", "ggplot2", "Hmisc", "viridis", "raster", "RColorBrewer", "cowplot", "rasterVis")
source("../src/ipak.R")
ipak(packages)
```

# stacked permutation importance plot

note if you change mind and want jackknife then the code is in the older plots for papers rmd

first need to make the data plot-ready

```{r create permutation & jack data}
maxr_lst <- list.files("../output/maxent/models/test-comb/", pattern = "maxentresults.csv", full.names = TRUE)
perm <- data.frame(matrix(ncol = 10, nrow = length(maxr_lst)))
colnames(perm) <- c("Month", "amo_sample", "amo_winter", "chl_surface", "nao_prev", "nao_sample", "nao_winter", "o2_depth", "salinity_depth", "temp_depth")
# jack <- data.frame(matrix(ncol = 10, nrow = length(maxr_lst)))
# colnames(jack) <- c("Month", "amo_sample", "amo_winter", "chl_surface", "nao_prev", "nao_sample", "nao_winter", "o2_depth", "salinity_depth", "temp_depth")

for (m in 1:length(maxr_lst)){
  maxr <- read.csv(maxr_lst[m])
  mth <- str_sub(maxr_lst[m], start = 35, end = 37) #selects the month from the filename
  perm$Month[m] <- mth
  # jack$Month[m] <- mth
  maxt <- tail(maxr, 1)
  perm$amo_sample[m] <- maxt$amo_sample.permutation.importance
  perm$amo_winter[m] <- maxt$amo_winter.permutation.importance
  perm$chl_surface[m] <- maxt$chl_surface.permutation.importance
  perm$nao_prev[m] <- maxt$nao_prev.permutation.importance
  perm$nao_sample[m] <- maxt$nao_sample.permutation.importance
  perm$nao_winter[m] <- maxt$nao_winter.permutation.importance
  perm$o2_depth[m] <- maxt$o2_depth.permutation.importance
  perm$salinity_depth[m] <- maxt$salinity_depth.permutation.importance
  perm$temp_depth[m] <- maxt$temp_depth.permutation.importance

  # jack$amo_sample[m] <- maxt$Training.gain.without.amo_sample
  # jack$amo_winter[m] <- maxt$Training.gain.without.amo_winter
  # jack$chl_surface[m] <- maxt$Training.gain.without.chl_surface
  # jack$nao_prev[m] <- maxt$Training.gain.without.nao_prev
  # jack$nao_sample[m] <- maxt$Training.gain.without.nao_sample
  # jack$nao_winter[m] <- maxt$Training.gain.without.nao_winter
  # jack$o2_depth[m] <- maxt$Training.gain.without.o2_depth
  # jack$salinity_depth[m] <- maxt$Training.gain.without.salinity_depth
  # jack$temp_depth[m] <- maxt$Training.gain.without.temp_depth
}

mthord <- c("mar", "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec") # a list of months to order by
permu <- perm[match(mthord, perm$Month), ] #reorder based on mthord
# jackk <- jack[match(mthord, jack$Month), ] #reorder based on mthord

# rename the months so they are capitalised

permu$Month <- capitalize(permu$Month)
# jackk$Month <- capitalize(jackk$Month)

@@

# colnames(jackk)[colnames(jackk)=="amo_sample"] <- "AMO_Sample"
# colnames(jackk)[colnames(jackk)=="amo_winter"] <- "AMO_Winter"
# colnames(jackk)[colnames(jackk)=="chl_surface"] <- "Chl_Surface"
# colnames(jackk)[colnames(jackk)=="nao_prev"] <- "NAO_Prev"
# colnames(jackk)[colnames(jackk)=="nao_sample"] <- "NAO_Sample"
# colnames(jackk)[colnames(jackk)=="nao_winter"] <- "NAO_Winter"
# colnames(jackk)[colnames(jackk)=="o2_depth"] <- "O2_Depth"
# colnames(jackk)[colnames(jackk)=="salinity_depth"] <- "Salinity_Depth"
# colnames(jackk)[colnames(jackk)=="temp_depth"] <- "Temp_Depth"

write.csv(permu, "../output/maxent/models/test-comb/permutation_avg.csv", row.names = FALSE)
# write.csv(jackk, "../output/maxent/models/test-comb/jack_avg.csv", row.names = FALSE)

@@

now make the plot

```{r permulation plot}
library(cowplot)
permu <- read.csv("../output/maxent/models/test-comb/permutation_avg.csv")

@@

permelt <- melt(permu, id = "Month") #need to melt the data into this format for ggplot
permelt$variable <- factor(permelt$variable, levels = c("AMO_Sample", "AMO_Winter", "NAO_Sample", "NAO_Prev", "NAO_Winter", "Chl_Surface", "O2_Depth", "Salinity_Depth", "Temp_Depth")) #reorder the variables
perpl <- ggplot(data = permelt, aes(x = Month, y = value, fill = variable)) + geom_bar(stat = "identity") +  scale_fill_viridis_d(direction = 1, labels = c("AMO Sample", "AMO Winter", "NAO Sample", "NAO Prev", "NAO Winter", "Chl Surface", "O2 Depth", "Salinity Depth", "Temp Depth")) + theme_half_open(12) + scale_x_discrete(limits = mthna) + labs(y = "Permutation Importance (%)")
png("../output/maxent/paper_plots/permutation_plot.png")
print(perpl)
dev.off()
```

histogram of probability by depth - by proportion

#max kappa thresholds

think about how to get proportions...

```{r}
mthord <- c("mar",  "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec")
cumst1 <- read.csv("../output/maxent/paper_plots/cellsperlayer.csv", header = TRUE)

```
based on Remi's code
```{r}
require(data.table)
require(purrr)
require(lubridate)
require(tidyverse)

@@

@@

ggsave("../output/maxent/paper_plots/test-overview.png", width = 210, height = 297, units = "mm", dpi = 600)
```

@@

@@

@@

get data to plot maxkap binary by depth... 
```{r maxkap by depth}
mthord <- c("mar", "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec")

for (m in 1:length(mthord)){
  mth <- mthord[m]
  ascdepth <- list.files(paste0("../output/maxent/paper_plots/prob_by_depth"), pattern = paste0("^", mth, "_maxkap.*asc"), full.names = TRUE)
  binaryavgstack <- stack(ascdepth)
  stackstathigh <- data.frame(matrix(ncol = 2, nrow = nlayers(binaryavgstack)))
  stackstathigh[ ,1] <- c(1:47)
  colnames(stackstathigh) <- c("asclayer", "prob_bin")

  for (lyr in 1:nlayers(binaryavgstack)){
    ascname <-  str_sub(ascdepth[lyr], 67, -5)
    dlay <- binaryavgstack[[lyr]]
    dlay <- calc(dlay, fun = binprob) #creates raster layer with cell values based on the binprob function
    lyrsum <- cellStats(dlay, sum)
    stackstathigh$prob_bin[lyr] <- lyrsum
    stackstathigh$asclayer[lyr] <- ascname

  }
  write.csv(stackstathigh, paste0("../output/maxent/paper_plots/prob_by_depth/", mth, "_maxkap_cellcount_bydepth_avg.csv"), row.names = FALSE)
}

think about how to get proportions...

@@

based on Remi's code - maxkappa derrived habitat by depth
```{r maxkappa habitat by depth}
require(data.table)
require(purrr)
require(lubridate)
require(tidyverse)
require(viridis)

@@

#get max and min depth layers based on max kappa
```{r maxkap min max depth}
kapd <- list.files("../output/maxent/paper_plots/prob_by_depth/" , pattern = "_maxkap_dis_prob_cells.csv", full.names = TRUE)
kapmmd <- data.frame(mth = character(10), min = numeric(10), max = numeric(10), range = numeric(10))
kapmmd$mth <- as.character(kapmmd$mth)

@@

#get lat long position of most northely, southerly, western, and eastern cell centroid (max kap)

```{r maxkap cell centroid}
filelist <- list.files("../output/maxent/paper_plots/hotspot_mapping/", pattern = "maxkap_single.asc", full.names = TRUE)
lonlat <- data.frame(mth = character(10), west = numeric(10), east = numeric(10), north = numeric(10), south = numeric(10))
lonlat$mth <- as.character(lonlat$mth)

@@

#get lat long position of most northely, southerly, western, and eastern cell centroid for oct, june, aug, may (to remove northern spots)

```{r}
filelist <- list.files("../output/maxent/paper_plots/hotspot_mapping/", pattern = "maxkap_single.asc", full.names = TRUE)
lonlat <- data.frame(mth = character(10), west = numeric(10), east = numeric(10), north = numeric(10), south = numeric(10))
lonlat$mth <- as.character(lonlat$mth)

@@

#background point selection plots

```{r}
test_back_df <- read.csv("../data/env/background_point_check/201406_allbackgroundpoints.csv", header = TRUE)
```

ok now create the first lot of random for 2014_06

```{r}
testbk10000 <- test_back_df[sample(nrow(test_back_df), 10000), ]  #where 10000 = number of rows to sample (large sample as per maxent)
testbk20000 <- test_back_df[sample(nrow(test_back_df), 20000), ]  
testbk50000 <- test_back_df[sample(nrow(test_back_df), 50000), ]  
testbk100000 <- test_back_df[sample(nrow(test_back_df), 100000), ]  
testbk190000 <- test_back_df[sample(nrow(test_back_df), 190000), ]  
```

make long data
```{r}
bkpt <- data.frame(no_points = character(370000), variable = character(370000), value = numeric(370000))
bkpt$variable <- as.character(bkpt$variable)
bkpt$no_points <- as.character(bkpt$no_points)

bkpt$no_points[1:10000] <- "10,000"
bkpt$no_points[10001:30000] <- "20,000"
bkpt$no_points[30001:80000] <- "50,000"
bkpt$no_points[80001:180000] <- "100,000"
bkpt$no_points[180001:370000] <- "190,000"
bkpt$variable <- "temp"
bkpt$value[1:10000] <- testbk10000$temp_depth
bkpt$value[10001:30000] <- testbk20000$temp_depth
bkpt$value[30001:80000] <- testbk50000$temp_depth
bkpt$value[80001:180000] <- testbk100000$temp_depth
bkpt$value[180001:370000] <- testbk190000$temp_depth

bkpt2 <- data.frame(no_points = numeric(370000), variable = character(370000), value = numeric(370000))
bkpt2$variable <- as.character(bkpt2$variable)
bkpt2$no_points[1:10000] <- "10,000"
bkpt2$no_points[10001:30000] <- "20,000"
bkpt2$no_points[30001:80000] <- "50,000"
bkpt2$no_points[80001:180000] <- "100,000"
bkpt2$no_points[180001:370000] <- "190,000"
bkpt2$variable <- "salinity"
bkpt2$value[1:10000] <- testbk10000$salinity_depth
bkpt2$value[10001:30000] <- testbk20000$salinity_depth
bkpt2$value[30001:80000] <- testbk50000$salinity_depth
bkpt2$value[80001:180000] <- testbk100000$salinity_depth
bkpt2$value[180001:370000] <- testbk190000$salinity_depth

bkpt3 <- data.frame(no_points = numeric(370000), variable = character(370000), value = numeric(370000))
bkpt3$variable <- as.character(bkpt3$variable)
bkpt3$no_points[1:10000] <- "10,000"
bkpt3$no_points[10001:30000] <- "20,000"
bkpt3$no_points[30001:80000] <- "50,000"
bkpt3$no_points[80001:180000] <- "100,000"
bkpt3$no_points[180001:370000] <- "190,000"
bkpt3$variable <- "o2"
bkpt3$value[1:10000] <- testbk10000$o2_depth
bkpt3$value[10001:30000] <- testbk20000$o2_depth
bkpt3$value[30001:80000] <- testbk50000$o2_depth
bkpt3$value[80001:180000] <- testbk100000$o2_depth
bkpt3$value[180001:370000] <- testbk190000$o2_depth

bkpt4 <- data.frame(no_points = numeric(370000), variable = character(370000), value = numeric(370000))
bkpt4$variable <- as.character(bkpt4$variable)
bkpt4$no_points[1:10000] <- "10,000"
bkpt4$no_points[10001:30000] <- "20,000"
bkpt4$no_points[30001:80000] <- "50,000"
bkpt4$no_points[80001:180000] <- "100,000"
bkpt4$no_points[180001:370000] <- "190,000"
bkpt4$variable <- "chl"
bkpt4$value[1:10000] <- testbk10000$chl_surface
bkpt4$value[10001:30000] <- testbk20000$chl_surface
bkpt4$value[30001:80000] <- testbk50000$chl_surface
bkpt4$value[80001:180000] <- testbk100000$chl_surface
bkpt4$value[180001:370000] <- testbk190000$chl_surface

bk <- rbind(bkpt, bkpt2, bkpt3, bkpt4)
bk$no_points <- as.factor(bk$no_points)
```

```{r}
ggplot(bk, aes(x = value)) +
    stat_density(aes(group = no_points, color = no_points),position="identity",geom="line") +
  facet_wrap(~variable,ncol=2)

problem here is that i don't want the axis shared

fiddle with the dataset again
```{r}
t <- subset(bk, variable == "temp")
t$no_points <- factor(t$no_points, levels = c("10,000", "20,000", "50,000", "100,000", "190,000"))
s <- subset(bk, variable == "salinity")
s$no_points <- factor(s$no_points, levels = c("10,000", "20,000", "50,000", "100,000", "190,000"))
o <- subset(bk, variable == "o2")
o$no_points <- factor(o$no_points, levels = c("10,000", "20,000", "50,000", "100,000", "190,000"))
c <- subset(bk, variable == "chl")
c$no_points <- factor(c$no_points, levels = c("10,000", "20,000", "50,000", "100,000", "190,000"))

plot
```{r}
p1 <- ggplot(t, aes(x = value)) + stat_density(aes(group = no_points, color = no_points),position="identity",geom="line", size=0.75) +theme_classic(base_size = 18) + labs(x = "Temperature (Kelvin)", y = "Density") + guides(color = guide_legend(title = "No Points"))
p2 <- ggplot(s, aes(x = value)) + stat_density(aes(group = no_points, color = no_points),position="identity",geom="line", size=0.75) + theme_classic(base_size = 18) + labs (x = "Salinity (PSU)", y = "Density") + guides(color = guide_legend(title = "No Points"))
p3 <- ggplot(o, aes(x = value)) + stat_density(aes(group = no_points, color = no_points),position="identity",geom="line", size=0.75) +theme_classic(base_size = 18) + labs (x = "Dissolved Oxygen (mmol\u207B\u00B3)", y = "Density") + guides(color = guide_legend(title = "No Points"))
p4 <- ggplot(c, aes(x = value)) + stat_density(aes(group = no_points, color = no_points),position="identity",geom="line", size=0.75) + theme_classic(base_size = 18) + labs (x = "Chlorophyll \u03b1  (mmol\u207B\u00B3)", y = "Density") + guides(color = guide_legend(title = "No Points"))

p <- ggarrange(p1, p2, p3, p4, ncol = 2, nrow = 2, common.legend = TRUE, legend = "bottom")

ggsave("../output/maxent/paper_plots/bkpt_201406.png", width = 210, height = 297, units = "mm", dpi = 600)

200710

```{r}
test_back_df <- read.csv("../data/env/background_point_check/200710_allbackgroundpoints.csv", header = TRUE)
```

ok now create the first lot of random for 200710

```{r}
testbk10000 <- test_back_df[sample(nrow(test_back_df), 10000), ]  #where 10000 = number of rows to sample (large sample as per maxent)
testbk20000 <- test_back_df[sample(nrow(test_back_df), 20000), ]  
testbk50000 <- test_back_df[sample(nrow(test_back_df), 50000), ]  
testbk100000 <- test_back_df[sample(nrow(test_back_df), 100000), ]  
testbk190000 <- test_back_df[sample(nrow(test_back_df), 190000), ]  
```

make long data
```{r}
bkpt <- data.frame(no_points = character(370000), variable = character(370000), value = numeric(370000))
bkpt$variable <- as.character(bkpt$variable)
bkpt$no_points <- as.character(bkpt$no_points)

bkpt$no_points[1:10000] <- "10,000"
bkpt$no_points[10001:30000] <- "20,000"
bkpt$no_points[30001:80000] <- "50,000"
bkpt$no_points[80001:180000] <- "100,000"
bkpt$no_points[180001:370000] <- "190,000"
bkpt$variable <- "temp"
bkpt$value[1:10000] <- testbk10000$temp_depth
bkpt$value[10001:30000] <- testbk20000$temp_depth
bkpt$value[30001:80000] <- testbk50000$temp_depth
bkpt$value[80001:180000] <- testbk100000$temp_depth
bkpt$value[180001:370000] <- testbk190000$temp_depth

bkpt2 <- data.frame(no_points = numeric(370000), variable = character(370000), value = numeric(370000))
bkpt2$variable <- as.character(bkpt2$variable)
bkpt2$no_points[1:10000] <- "10,000"
bkpt2$no_points[10001:30000] <- "20,000"
bkpt2$no_points[30001:80000] <- "50,000"
bkpt2$no_points[80001:180000] <- "100,000"
bkpt2$no_points[180001:370000] <- "190,000"
bkpt2$variable <- "salinity"
bkpt2$value[1:10000] <- testbk10000$salinity_depth
bkpt2$value[10001:30000] <- testbk20000$salinity_depth
bkpt2$value[30001:80000] <- testbk50000$salinity_depth
bkpt2$value[80001:180000] <- testbk100000$salinity_depth
bkpt2$value[180001:370000] <- testbk190000$salinity_depth

bkpt3 <- data.frame(no_points = numeric(370000), variable = character(370000), value = numeric(370000))
bkpt3$variable <- as.character(bkpt3$variable)
bkpt3$no_points[1:10000] <- "10,000"
bkpt3$no_points[10001:30000] <- "20,000"
bkpt3$no_points[30001:80000] <- "50,000"
bkpt3$no_points[80001:180000] <- "100,000"
bkpt3$no_points[180001:370000] <- "190,000"
bkpt3$variable <- "o2"
bkpt3$value[1:10000] <- testbk10000$o2_depth
bkpt3$value[10001:30000] <- testbk20000$o2_depth
bkpt3$value[30001:80000] <- testbk50000$o2_depth
bkpt3$value[80001:180000] <- testbk100000$o2_depth
bkpt3$value[180001:370000] <- testbk190000$o2_depth

bkpt4 <- data.frame(no_points = numeric(370000), variable = character(370000), value = numeric(370000))
bkpt4$variable <- as.character(bkpt4$variable)
bkpt4$no_points[1:10000] <- "10,000"
bkpt4$no_points[10001:30000] <- "20,000"
bkpt4$no_points[30001:80000] <- "50,000"
bkpt4$no_points[80001:180000] <- "100,000"
bkpt4$no_points[180001:370000] <- "190,000"
bkpt4$variable <- "chl"
bkpt4$value[1:10000] <- testbk10000$chl_surface
bkpt4$value[10001:30000] <- testbk20000$chl_surface
bkpt4$value[30001:80000] <- testbk50000$chl_surface
bkpt4$value[80001:180000] <- testbk100000$chl_surface
bkpt4$value[180001:370000] <- testbk190000$chl_surface

bk <- rbind(bkpt, bkpt2, bkpt3, bkpt4)
bk$no_points <- as.factor(bk$no_points)
```

fiddle with the dataset again
```{r}
t <- subset(bk, variable == "temp")
t$no_points <- factor(t$no_points, levels = c("10,000", "20,000", "50,000", "100,000", "190,000"))
s <- subset(bk, variable == "salinity")
s$no_points <- factor(s$no_points, levels = c("10,000", "20,000", "50,000", "100,000", "190,000"))
o <- subset(bk, variable == "o2")
o$no_points <- factor(o$no_points, levels = c("10,000", "20,000", "50,000", "100,000", "190,000"))
c <- subset(bk, variable == "chl")
c$no_points <- factor(c$no_points, levels = c("10,000", "20,000", "50,000", "100,000", "190,000"))

plot
```{r}
p1 <- ggplot(t, aes(x = value)) + stat_density(aes(group = no_points, color = no_points),position="identity",geom="line", size=0.75) +theme_classic(base_size = 18) + labs(x = "Temperature (Kelvin)", y = "Density") + guides(color = guide_legend(title = "No Points"))
p2 <- ggplot(s, aes(x = value)) + stat_density(aes(group = no_points, color = no_points),position="identity",geom="line", size=0.75) + theme_classic(base_size = 18) + labs (x = "Salinity (PSU)", y = "Density") + guides(color = guide_legend(title = "No Points"))
p3 <- ggplot(o, aes(x = value)) + stat_density(aes(group = no_points, color = no_points),position="identity",geom="line", size=0.75) +theme_classic(base_size = 18) + labs (x = "Dissolved Oxygen (mmol\u207B\u00B3)", y = "Density") + guides(color = guide_legend(title = "No Points"))
p4 <- ggplot(c, aes(x = value)) + stat_density(aes(group = no_points, color = no_points),position="identity",geom="line", size=0.75) + theme_classic(base_size = 18) + labs (x = "Chlorophyll \u03b1  (mmol\u207B\u00B3)", y = "Density") + guides(color = guide_legend(title = "No Points"))

p <- ggarrange(p1, p2, p3, p4, ncol = 2, nrow = 2, common.legend = TRUE, legend = "bottom")

ggsave("../output/maxent/paper_plots/bkpt_200710.png", width = 210, height = 297, units = "mm", dpi = 600)

>>>>

---
title: "maxent model plus plot creation"
author: "Samantha Andrews"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

# Overview
A note to anyone who might happen to stumble across this... I am a beginner in R and have had no exposure to similar languages. I don't know what I'm doing. The code herein is unlikely to be elegant and there area probably more efficient ways of running the code.

Built with 'r getRversion()'..

# Package dependencies
You can install and load them using the following code which uses a function called [ipak](https://gist.github.com/stevenworthington/3178163). Note this function checks to see if the packages are installed first before loading.

```{r pre-install & load packages, include=FALSE}
options(java.parameters = "-Xmx3g")#needs to run before dismo loaded - boosts memory for java to 3g
packages <- c("dismo", "stringr", "rJava", "raster", "ecospat", "bossMaps", "lattice", "enmSdm", "ggplot2", "plyr", "reshape2", "tidyverse", "patchwork", "purrr", "Hmisc", "cowplot", "sf", "lubridate")
source("../src/ipak.R")
ipak(packages)
```

load the presence-background dataset

```{r load pres-background dataset}
presback <- read.csv("../output/bio/presab_nafo_yr.csv", header = TRUE) # this file is available in the Figshare project
```

For each monthly model, subset the data to the month you want, and the variables you want (Based on spearmans and VIF)
Ex: edit this list for variables used in each month (or simply replace with a single list if, as in this case, months have the same variables)

@@

```{r train and test inc AIC for reg val}
presyrs <-  #this is to get years with complete presence data

modelmonth <- cbind(paste0(mthna ,unique(prback$year)))
aicetc <- read.csv("../data/env/AICcCalculations.csv", header = TRUE) #pre-created csv file
betavallist <- cbind(aicetc$regularization)

for (year in 1:nrow(modelmonth)){
  timeslice <- modelmonth[year] #testing timeslice
  yr <- str_sub(timeslice, -4) #the year being used for testing
  train <- subset(prback, year != yr) #training data is all years but testing
  trainpa <- train$occurrence #creates a vector of presence/background ID (1/0)
  test <-  subset(prback, year == yr) #creates the testing dataset
  testp <- subset(test, occurrence == "1") #splits the testing dataset into presences
  testp <- subset(testp, select = -c(year, occurrence)) #don't need year and occurrence for modelling
  testb <- subset(test, occurrence == "0") #splits the testing dataset into background
  testb <- subset(testb, select = -c(year, occurrence)) #don't need year and occurrence for modelling
  train  <- subset(train, select = -c(year, occurrence)) #don't need year and occurrence for modelling

  for (h in 1:nrow(betavallist)){
    betaval <-  betavallist[h]
    mod <- maxent(x = train, p = trainpa, path = paste0("../output/maxent/models/", timeslice), args=c('plots=FALSE', "maximumiterations=5000", "responsecurves=TRUE", "jackknife=TRUE", "threads=8", "addsamplestobackground=TRUE", paste0("betamultiplier=", betavallist[h])))

    #count number of non-zero coefficients, excluding the final four rows
    coefile <- read.csv(paste0("../output/maxent/models/", timeslice, "/species.lambdas"), header = FALSE, stringsAsFactors=FALSE)
    coefile <- head(coefile, -4)
    coefile2 <- subset(coefile, V2 > 0 | V2 < 0)
    nocoeff <- nrow(coefile2)
    bick <- sum(as.numeric(coefile2[, 2])) #dodgy 
    bic <- (bick*log(sum(trainpa == 1)) - 2*11) #dodgy

    #no presences used in model (less those used for testing)
    maxresults <- read.csv(paste0("../output/maxent/models/", timeslice, "/maxentResults.csv"), header = TRUE)
    notrainingpoints <- maxresults$X.Training.samples
    testauc <- maxresults$Test.AUC
    trainauc <- maxresults$Training.AUC

    #natural log
    samplepred <- read.csv(paste0("../output/maxent/models/", timeslice, "/species_samplePredictions.csv"), header = TRUE)
    samplepred$nat_log <- log(samplepred$Raw.prediction)
    sumsamplepred <- sum(samplepred$nat_log)

   #test auc scores and other lovely things you might want to include e.g.
    ev <- evaluate(p = testp, a= testb, model = mod)
    evauc <- ev@auc
    evcor <- ev@cor
    evpcor <- ev@pcor
    evnp <- ev@np
    evna <- ev@na
    evmaxtprtnr <- threshold(ev, 'spec_sens')
    evmaxkap <- threshold(ev, 'kappa')
    evno_omission <- threshold(ev, 'no_omission')
    evprevalence <- threshold(ev, 'prevalence')
    evequal_sens_spec <- threshold(ev, 'equal_sens_spec')
    evsensitivity <- threshold(ev, 'sensitivity')

    for (i in 1:nrow(aicetc)){
    if(aicetc$regularization[i] == betaval){
      aicetc$train_AUC[i] <- trainauc
      aicetc$no_trainng_points[i] <- notrainingpoints
      aicetc$coefficients[i] <- nocoeff
      aicetc$sum.of.logs[i] <- sumsamplepred
      aicetc$test_AUC[i] <- evauc
      aicetc$test_evcor[i] <- evcor
      aicetc$test_evpcor[i] <- evpcor
      aicetc$test_nopresence[i] <- evnp
      aicetc$test_nobackground[i] <- evna
      aicetc$test_maxsumsenspec[i] <- evmaxtprtnr
      aicetc$test_maxkap[i] <- evmaxkap
      aicetc$test_no_omission[i] <- evno_omission
      aicetc$test_prevalence[i] <- evprevalence
      aicetc$test_equal_sens_spec[i] <- evequal_sens_spec
      aicetc$test_sensitivity[i] <- evsensitivity
      aicetc$test_bic[i] <- bic #not sure this is correct...
      aicetc$beta[i] <- betaval
      aicetc$mpa_0.9_cum[i] <- ecospat.mpa(samplepred$Cumulative.prediction, perc = 0.9) #90%

    aicetc$AIC[i] <- ((2*aicetc$coefficients[i])-(2*aicetc$sum.of.logs[i]))
    aicetc$AICc[i] <- (-2*(aicetc$sum.of.logs[i])+((2*aicetc$coefficients[i])*(aicetc$no_trainng_points[i]/(aicetc$no_trainng_points[i] - aicetc$coefficients[i]-1))))
    }
   }
  }

write.csv(aicetc, paste0("../output/maxent/models/", timeslice, "/AICforMaxentReg.csv"), row.names = FALSE)

AICc selection
select the models you want for each year and run those models again (primarily based on lowest AICc but check test AUC and other things too)

@@

# average the maxent results
did this after splitting layers back into their individual depths as per Bentlage 2013 https://doi.org/10.1111/geb.12049

@@

#the following cde was manually edited for each month

@@

@@

# stacked permutation importance plot
This is Figure 2

```{r create permutation & jack data}
maxr_lst <- list.files("../output/maxent/models/test-comb/", pattern = "maxentresults.csv", full.names = TRUE)
perm <- data.frame(matrix(ncol = 10, nrow = length(maxr_lst)))
colnames(perm) <- c("Month", "amo_sample", "amo_winter", "chl_surface", "nao_prev", "nao_sample", "nao_winter", "o2_depth", "salinity_depth", "temp_depth")

for (m in 1:length(maxr_lst)){
  maxr <- read.csv(maxr_lst[m])
  mth <- str_sub(maxr_lst[m], start = 35, end = 37) #selects the month from the filename
  perm$Month[m] <- mth
  maxt <- tail(maxr, 1)
  perm$amo_sample[m] <- maxt$amo_sample.permutation.importance
  perm$amo_winter[m] <- maxt$amo_winter.permutation.importance
  perm$chl_surface[m] <- maxt$chl_surface.permutation.importance
  perm$nao_prev[m] <- maxt$nao_prev.permutation.importance
  perm$nao_sample[m] <- maxt$nao_sample.permutation.importance
  perm$nao_winter[m] <- maxt$nao_winter.permutation.importance
  perm$o2_depth[m] <- maxt$o2_depth.permutation.importance
  perm$salinity_depth[m] <- maxt$salinity_depth.permutation.importance
  perm$temp_depth[m] <- maxt$temp_depth.permutation.importance

mthord <- c("mar", "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec") # a list of months to order by
permu <- perm[match(mthord, perm$Month), ] #reorder based on mthord

# rename the months so they are capitalised
permu$Month <- capitalize(permu$Month)

@@

write.csv(permu, "../output/maxent/models/test-comb/permutation_avg.csv", row.names = FALSE)

@@

now make the plot
```{r permulation plot}
permu <- read.csv("../output/maxent/models/test-comb/permutation_avg.csv")

@@

perm2 <- melt(permu, id = "Month") #need to melt the data into this format for ggplot
perm2$variable <- factor(perm2$variable, levels = c("AMO_Sample", "AMO_Winter", "NAO_Sample", "NAO_Prev", "NAO_Winter", "Chl_Surface", "O2_Depth", "Salinity_Depth", "Temp_Depth")) #reorder the variables
perpl <- ggplot(data = perm2, aes(x = Month, y = value, fill = variable)) + geom_bar(stat = "identity") +  scale_fill_viridis_d(direction = 1, labels = c("AMO Sample", "AMO Winter", "NAO Sample", "NAO Prev", "NAO Winter", "Chl Surface", "O2 Depth", "Salinity Depth", "Temp Depth")) + theme_half_open(12) + scale_x_discrete(limits = mthna) + labs(y = "Permutation Importance (%)")
png("../output/maxent/paper_plots/permutation_plot.png")
print(perpl)
dev.off()
```

@@

This is Figure 3

Code base provided by Rémi Daigle (https://remidaigle.wordpress.com/)

@@

ggsave("../output/maxent/paper_plots/prob-overview.png", width = 210, height = 297, units = "mm", dpi = 600)
```

This is figure 4

Get proprotions
```{r}
mthord <- c("mar",  "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec")
cumst1 <- read.csv("../output/maxent/paper_plots/cellsperlayer.csv", header = TRUE)

plot - Code base provided by Rémi Daigle (https://remidaigle.wordpress.com/)

```{r}
require(data.table)
require(purrr)
require(lubridate)
require(tidyverse)

@@

Get max kappa thresholds

@@

plot - Code base provided by Rémi Daigle (https://remidaigle.wordpress.com/)
This is Figure 5

@@

get data to plot maxkap binary by depth plot (Figure 6)

Get proportions...

@@

plot - Code base provided by Rémi Daigle (https://remidaigle.wordpress.com/)
This is Figure 6

```{r maxkappa habitat by depth}
require(data.table)
require(purrr)
require(lubridate)
require(tidyverse)
require(viridis)

@@

#stats reported in the manuscript

get max and min depth layers based on max kappa
```{r maxkap min max depth}
kapd <- list.files("../output/maxent/paper_plots/prob_by_depth/" , pattern = "_maxkap_dis_prob_cells.csv", full.names = TRUE)
kapmmd <- data.frame(mth = character(10), min = numeric(10), max = numeric(10), range = numeric(10))
kapmmd$mth <- as.character(kapmmd$mth)

@@

get lat long position of most northely, southerly, western, and eastern cell centroid (max kap)
```{r maxkap cell centroid}
filelist <- list.files("../output/maxent/paper_plots/hotspot_mapping/", pattern = "maxkap_single.asc", full.names = TRUE)
lonlat <- data.frame(mth = character(10), west = numeric(10), east = numeric(10), north = numeric(10), south = numeric(10))
lonlat$mth <- as.character(lonlat$mth)

@@

get lat long position of most northely, southerly, western, and eastern cell centroid for oct, june, aug, may (to remove northern spots)
```{r}
filelist <- list.files("../output/maxent/paper_plots/hotspot_mapping/", pattern = "maxkap_single.asc", full.names = TRUE)
lonlat <- data.frame(mth = character(10), west = numeric(10), east = numeric(10), north = numeric(10), south = numeric(10))
lonlat$mth <- as.character(lonlat$mth)

@@
